Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-06-10 doc/Manuscript/article.tex
Updates from Overleaf

\begin{abstract}
This article presents novel applications of machine learning methods to the problem of event classification in an active target detector, the Active Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant.

@@

\begin{table}[hbtp]
\centering
\caption{Descriptions of number of events in the data. In principle we can simulate an arbitrary amount of data, but the following quantities were shown to be sufficient for these studies.{\color{blue} reference results?}}\label{tab:datasets}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Total &  $8000$ & $51891$ & $49169$ \\
labelled & $2400$ & $1774$ &  $1582$ \\ 
\bottomrule
\end{tabular}
\end{table}

@@

In contrast with the classification work of \citet{Kuchera2019}, we do not assume access to ground truth labels and are trying to solve a fundamentally different learning problem. Thus, rather than fine-tuning a pre-trained network under the supervised learning regime, we extract the output of the pre-trained network's last convolutional layer as a latent representation of the events on which we will apply a clustering algorithm.

The clustering algorithm we found to give the strongest result is the K-means algorithm. This algorithm is contingent on having a distance measurement between two datapoints. We found that the standard choice of a euclidian norm on the vector differences provided the best outcomes.

@@

As an alternative to using a pre-trained model in conjunction with a clustering algorithm, we have also considered bespoke end-to-end clustering models. These models are optimized specifically to the AT-TPC data with the objective of clustering. For this paper, we considered the deep convolutional embedded clustering (DCEC) by \citet{Guo2017} as well as the mixture of autoencoders (MIXAE) method introduced by \citet{Zhang}. While we were able to reproduce the authors' results on their data, the DCEC algorithm proved unable to cluster AT-TPC in our implementation. However, this provides valuable insight as the seeds of the clusters are constructed by a K-means algorithm. This insight contrasts with our positive results from applying a pre-trained model with k-means and highlights potentially significant differences in models trained on a supervised or unsupervised objective for clustering tasks in nuclear physics.

@@

\begin{figure}[tb]
	\centering
	\includegraphics[width=.8\textwidth]{plots/mixae.pdf}
	\caption[Mixture of autoencoders schematic]{Schematic of a MIXAE model. A sample $\hat{\boldsymbol{x}}$ is compressed to set of lower-dimensional representations $\{\boldsymbol{z}^{(i)}\}$ by $N$ autoencoders. These samples are concatenated and passed through an auxiliary assignment network that predicts a confidence of cluster belonging for each autoencoder. For further details see the text. Figure adapted from \citet{Zhang}}
	\label{fig:mixae}
\end{figure}
{\color{blue} This section ends quite abruptly -MPK}
\section{Results and Discussions}\label{sec:results}
\begin{itemize}
\item Clear out the "methods" like parts.
\item merge 5.1 and 5.2 - keeping point 1 in mind.
\item merge 5.3 5.4 and 5.6 and create table from figure 10, 11 and 12. 
\item go over 5.7 - mention instability in MIXAE results, peak performance of MIXAE is good - consistency is not. 
\item remove fig. 14 and fig 15.
\end{itemize}

@@

>>>>

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant.

@@

\begin{table}[hbtp]
\centering
\caption{Descriptions of number of events in the data.}\label{tab:datasets}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Total &  $8000$ & $51891$ & $49169$ \\
labelled & $2400$ & $1774$ &  $1582$ \\ 
\bottomrule
\end{tabular}
\end{table}

@@

In contrast with the classification work of \citet{Kuchera2019}, we do not assume access to ground truth labels and are trying to solve a fundamentally different learning problem. Thus, rather than fine-tuning a pre-trained network under the supervised learning regime, we extract the output of the pre-trained network's last convolutional layer as a latent representation of the events. We then cluster events based on this representation using the K-means++ algorithm with the standard choice of Euclidean distance.

@@

As an alternative to using a pre-trained model in conjunction with a clustering algorithm, we have also considered bespoke end-to-end clustering models. These models are optimized specifically to the AT-TPC data with the objective of clustering. We considered the deep convolutional embedded clustering (DCEC) by \citet{Guo2017} as well as the mixture of autoencoders (MIXAE) method introduced by \citet{Zhang}. While we were able to reproduce the authors' results on their data, the DCEC algorithm proved unable to cluster AT-TPC in our implementation. However, this provides valuable insight as the seeds of the clusters are constructed by a K-means algorithm. This insight contrasts with our positive results from applying a pre-trained model with k-means and highlights potentially significant differences in models trained on a supervised or unsupervised objective for clustering tasks in nuclear physics.

@@

\begin{figure}[tb]
	\centering
	\includegraphics[width=.8\textwidth]{plots/mixae.pdf}
	\caption[Mixture of autoencoders schematic]{Schematic of a MIXAE model. A sample $\hat{\boldsymbol{x}}$ is compressed to set of lower-dimensional representations $\{\boldsymbol{z}^{(i)}\}$ by $N$ autoencoders. These samples are concatenated and passed through an auxiliary assignment network that predicts a confidence of cluster belonging for each autoencoder. For further details see the text. Figure adapted from \citet{Zhang}}
	\label{fig:mixae}
\end{figure}
{\color{blue} This section ends quite abruptly -MPK}

\section{Results and Discussions}\label{sec:results}
\begin{itemize}
\item Clear out the "methods" like parts.
\item merge 5.1 and 5.2 - keeping point 1 in mind.
\item merge 5.3 5.4 and 5.6 and create table from figure 10, 11 and 12. 
\item go over 5.7 - mention instability in MIXAE results, peak performance of MIXAE is good - consistency is not. 
\item remove fig. 14 and fig 15.
\end{itemize}

@@

From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. It is also clear that there is significant potential in both avenues explored in this work. 
In particular, the surprising ability of the K-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis. The consistency of the predictive quality of the K-means model also means that
